---
title: "PS270 Final Project"
author: "Neela Hammer"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts (milestone 1)

I am interested in exploring data related to how media consumption shapes voting habits, what countries are more prone to conflict, or something related to political nonprofits. 

## Research question (milestone 2)

My research question for this project is: Does change or decline in freedom scores between 2013 and 2025 differ by world region? My hypothesis is that Eurasia and the Middle East have experienced a larger decline in freedom than other regions due to war and occupation. My independent variable is world region, which is measured categorically by dividing countries into groups based on location. My dependent variable is freedom score as determined by Freedom House's Freedom in the World dataset. Each country is assigned to free, partially free, or not free status and is ranked from 0-4 on a set of freedom-related criteria. I will examine the changes in these rankings over time. If the data shows that there is a pattern of countries in Eurasia and/or the Middle East declining the most, my hypothesis will be correct. If another region has had larger declines, my hypothesis will be incorrect. 

```{r}
library(tidyverse)
FIWdata <-read_csv("data/FIWdatacsv.csv")
head(FIWdata)
```

## Data cleaning

```{r}
FIWdata <- FIWdata |>
  rename(
    country_territory = `Freedom in the World 2013-2025 Raw Data`,
    region = `...2`,
    ct = `...3`,
    edition = `...4`,
    status = `...5`,
    pr_rating = `...6`,
    cl_rating = `...7`,
    a1 = `...8`, a2 = `...9`, a3 = `...10`, a = `...11`,
    b1 = `...12`, b2 = `...13`, b3 = `...14`, b4 = `...15`, b = `...16`,
    c1 = `...17`, c2 = `...18`, c3 = `...19`, c = `...20`,
    add_q = `...21`, add_a = `...22`, pr = `...23`,
    d1 = `...24`, d2 = `...25`, d3 = `...26`, d4 = `...27`, d = `...28`,
    e1 = `...29`, e2 = `...30`, e3 = `...31`, e = `...32`,
    f1 = `...33`, f2 = `...34`, f3 = `...35`, f4 = `...36`, f = `...37`,
    g1 = `...38`, g2 = `...39`, g3 = `...40`, g4 = `...41`, g = `...42`,
    cl = `...43`, total = `...44`
    )
```

```{r}
FIWdata <- FIWdata[-1, ]
```

## Data visualization (milestone 3)

```{r}
region_time_plot <- FIWdata |>
  mutate(edition = as.numeric(edition), 
         total = as.numeric(total)) |>
  group_by(edition, region) |>
  summarize(mean_total = mean(total, na.rm = TRUE)) |>
  ggplot(aes(x = edition, y = mean_total, color = region)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2013,2025, by = 1)) +
  labs(
    title = "Freedom in world regions over time",
    x = "Year",
    y = "Total score on freedom-indicating variables",
    caption = "Data from Freedom in the World reports"
  )
region_time_plot
```

This scatterplot shows the relationship between region of the world and freedom score over time. Each line represents the mean freedom score of one world region at different points in the last 12 years. 

## Visualization and Analysis (milestone 4)

### Liner model for global freedom over time:

```{r}
FIW2 <- FIWdata |>
  mutate(edition = as.numeric(edition), 
         total = as.numeric(total)) |>
  group_by(edition, region) |>
  summarize(mean_total = mean(total))
```


```{r}
global_model <- lm(mean_total ~ edition, data = FIW2)
global_model
```

The intercept is not substantive here as it predicts the mean freedom score at year 0, which is not included in our dataset or important for the research question, which focuses on 2013-2025. 

The edition coefficient means that the average freedom score in the world as a whole over time decreases at an average of 0.41 points per year. This highlights that freedom in the world is overall declining. 

### Now, models isolating coefficients for each region:

```{r}
regional_model <- lm(mean_total ~ region * edition, data = FIW2)
regional_model
```

This model shows the effect of both year and region on mean total freedom score. The region coefficients alone represent the predicted change in freedom since year 0, which is not relevant to this situation. 

However, each region:edition coefficient represents the slope/change of freedom in a single world region from 2013-2025. The edition coefficient represents Africa, which this model predicts has declined 0.49 points on average per year. Each other coefficient represents the difference in slope from Africa, the base group.

According to this model, every region has a negative change in freedom during this time period. Here are the values of each slope:

Africa: -0.48518 points per year

Americas: -0.39027 points per year

Asia: -0.20086 points per year

Eurasia: -0.66443 points per year

Europe: -0.31957 points per year

Middle East: -0.41831 points per year

This means that Eurasia has the fastest decline in freedom over 2013-2025, and Asia has the slowest. According to this model, my hypothesis was partly correct. I predicted freedom in Eurasia and the Middle East had declined the most. In reality, Eurasia has declined the most, then Africa, then the Middle East. 

### Visualization of the linear models

```{r}
regional_tibble <- tibble(
  region = c("Eurasia", "Africa", "Middle East", "Americas", "Europe", "Asia"),
  slope = c(-0.66443, -0.48518, -0.41831, -0.39027, -0.31957, -0.20086)
)

regional_tibble |>
  mutate(slope_positive = abs(slope)) |>
  ggplot(aes(x = reorder(region, slope_positive), y = slope_positive)) +
  geom_col() +
  labs(
    title = "Slope of Freedom Decline in World Regions from 2013-2025",
    x = "Region",
    y = "Decline in freedom points",
    caption = "blue line represents slope of decline globally (average across all regions)"
  ) +
  geom_hline(yintercept = 0.4131, color = "blue")
```




