---
title: "Freedom in World Regions Over Time"
author: "Neela Hammer"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

### Research question
My research question for this project is: Does world region have an effect on the magnitude of change in freedom from 2013-2025?

### Hypothesizing:
My hypothesis is that **Eurasia** and the **Middle East** have experienced a larger decline in freedom than other regions due to war and occupation. Specifically, I think the wars in Ukraine and in Gaza could be causing a faster freedom decline. War can cause democratic decline as rulers are more autocratic, harsher policies are put in place to protect the country and people are subject to increased violence. 

### Importance
This topic is important because while all world regions have seen a freedom decline in the past 13 years, the effects can disproportionately affect specific regions and groups of people. It's important to understand how different areas of the world are changing and how ruling styles, conflicts or citizens' rights impact overall freedom. Examining this pattern will allow us to understand where democracy is weakening, where people are struggling to live freely and where intervention is most needed. With this information, international organizations, policymakers or other people in power could know where they most need to route their resources and energy to protect peoples' rights and freedoms. 

## Data

My data comes from the Freedom House's Freedom in the World dataset, which collects data on various aspects of freedom in different countries over different years. The data runs from 2013 to 2025, so this is the timeline that my analysis will focus on. The study used a **difference-in-differences** design because the data is collected from multiple countries over multiple years. This allows me to see the difference both among regions and across time. 

### Variables
My independent variable is **world region**, which is measured categorically by dividing countries into groups based on location. My dependent variable is **total freedom score** as determined by Freedom House's Freedom in the World dataset. Each country is assigned to free, partially free, or not free status and is ranked from 0-4 on a set of freedom-related criteria. I will examine the changes in these rankings, focusing on the total freedom score (sum of scores on all criteria), over time. 

### Cleaning the data
The code I wrote to clean the data is included in the code section at the bottom of the report. Mainly, I uploaded the dataset, renamed columns to make them clear and removed the first row which did not include any data. Throughout the process I also turned character strings into numeric values to be able to analyze means and other numerical statistics. 

```{r load data, include = FALSE}
library(tidyverse)
FIWdata <-read_csv("data/FIWdatacsv.csv")
head(FIWdata)
```


```{r clean columns, include = FALSE}
FIWdata <- FIWdata |>
  rename(
    country_territory = `Freedom in the World 2013-2025 Raw Data`,
    region = `...2`,
    ct = `...3`,
    edition = `...4`,
    status = `...5`,
    pr_rating = `...6`,
    cl_rating = `...7`,
    a1 = `...8`, a2 = `...9`, a3 = `...10`, a = `...11`,
    b1 = `...12`, b2 = `...13`, b3 = `...14`, b4 = `...15`, b = `...16`,
    c1 = `...17`, c2 = `...18`, c3 = `...19`, c = `...20`,
    add_q = `...21`, add_a = `...22`, pr = `...23`,
    d1 = `...24`, d2 = `...25`, d3 = `...26`, d4 = `...27`, d = `...28`,
    e1 = `...29`, e2 = `...30`, e3 = `...31`, e = `...32`,
    f1 = `...33`, f2 = `...34`, f3 = `...35`, f4 = `...36`, f = `...37`,
    g1 = `...38`, g2 = `...39`, g3 = `...40`, g4 = `...41`, g = `...42`,
    cl = `...43`, total = `...44`
    )
```

```{r remove row, include = FALSE}
FIWdata <- FIWdata[-1, ]
```

### Criteria for hypothesis
To test my hypothesis, I will run a linear regression for the interaction of region and year on total freedom score. I would expect to see Eurasia or the Middle East having the steepest negative slope. If the slope is positive or another region has a steeper decline, I will know the hypothesis is incorrect.

I will find the p-value for this coeffecient to statistically determine which hypothesis is correct:
**Null:** Region of the world has no effect on change in freedom score over time.
**Alternative:** Region of the world DOES have an effect on change in freedom score over time. The ATE does not equal 0.
If the p-value is less than or equal to 0.05 I will reject the null. If it is greater, I will fail to reject the null hypothesis of no difference. 

### Visualization of freedom over time by region
This plot shows the relationship between region of the world and freedom score over time. Each line represents the mean freedom score of one world region at different points in the last 12 years. 

```{r region time plot, echo = FALSE}
region_time_plot <- FIWdata |>
  mutate(edition = as.numeric(edition), 
         total = as.numeric(total)) |>
  group_by(edition, region) |>
  summarize(mean_total = mean(total, na.rm = TRUE)) |>
  ggplot(aes(x = edition, y = mean_total, color = region)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2013,2025, by = 1)) +
  labs(
    title = "Freedom in world regions over time",
    x = "Year",
    y = "Total score on freedom-indicating variables",
    caption = "Data from Freedom in the World reports"
  )
region_time_plot
```


## Results

### Liner model for global freedom over time:

```{r}
FIW2 <- FIWdata |>
  mutate(edition = as.numeric(edition), 
         total = as.numeric(total)) |>
  group_by(edition, region) |>
  summarize(mean_total = mean(total))
```


```{r}
global_model <- lm(mean_total ~ edition, data = FIW2)
global_model

global_tibble <- tibble(
  intercept = 886.0584,
  edition = -0.4131
)
knitr::kable(global_tibble)
```

The intercept is not substantive here as it predicts the mean freedom score at year 0, which is not included in our dataset or important for the research question, which focuses on 2013-2025. 

The edition coefficient means that the average freedom score in the world as a whole over time decreases at an average of 0.41 points per year. This highlights that freedom in the world is overall declining. 

### Now, models isolating coefficients for each region:

```{r}
regional_model <- lm(mean_total ~ region * edition, data = FIW2)

regional_tibble <- tibble(
  region = c("Eurasia", "Africa", "Middle East", "Americas", "Europe", "Asia"),
  slope = c(-0.66443, -0.48518, -0.41831, -0.39027, -0.31957, -0.20086)
)
knitr::kable(regional_tibble)
```

This model shows the effect of both year and region on mean total freedom score. The region coefficients alone represent the predicted change in freedom since year 0, which is not relevant to this situation. 

However, each region:edition coefficient represents the slope/change of freedom in a single world region from 2013-2025. The edition coefficient represents Africa, which this model predicts has declined 0.49 points on average per year. Each other coefficient represents the difference in slope from Africa, the base group.

According to this model, every region has a negative change in freedom during this time period. Here are the values of each slope:

Africa: -0.48518 points per year

Americas: -0.39027 points per year

Asia: -0.20086 points per year

Eurasia: -0.66443 points per year

Europe: -0.31957 points per year

Middle East: -0.41831 points per year

This means that Eurasia has the fastest decline in freedom over 2013-2025, and Asia has the slowest. According to this model, my hypothesis was partly correct. I predicted freedom in Eurasia and the Middle East had declined the most. In reality, Eurasia has declined the most, then Africa, then the Middle East. 

### Visualization of the linear models

```{r}
regional_tibble <- tibble(
  region = c("Eurasia", "Africa", "Middle East", "Americas", "Europe", "Asia"),
  slope = c(-0.66443, -0.48518, -0.41831, -0.39027, -0.31957, -0.20086)
)

regional_tibble |>
  mutate(slope_positive = abs(slope)) |>
  ggplot(aes(x = reorder(region, slope_positive), y = slope_positive)) +
  geom_col() +
  labs(
    title = "Slope of Freedom Decline in World Regions from 2013-2025",
    x = "Region",
    y = "Decline in freedom points",
    caption = "blue line represents slope of decline globally (average across all regions)"
  ) +
  geom_hline(yintercept = 0.4131, color = "blue")
```




