---
title: "PS270 Final Project"
author: "Neela Hammer"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to how media consumption shapes voting habits, what countries are more prone to conflict, or something related to political nonprofits. 

## Research question

My research question for this project is: Does change or decline in freedom scores between 2013 and 2025 differ by world region? My hypothesis is that Eurasia and the Middle East have experienced a larger decline in freedom than other regions due to war and occupation. My independent variable is world region, which is measured categorically by dividing countries into groups based on location. My dependent variable is freedom score as determined by Freedom House's Freedom in the World dataset. Each country is assigned to free, partially free, or not free status and is ranked from 0-4 on a set of freedom-related criteria. I will examine the changes in these rankings over time. If the data shows that there is a pattern of countries in Eurasia and/or the Middle East declining the most, my hypothesis will be correct. If another region has had larger declines, my hypothesis will be incorrect. 

```{r}
library(tidyverse)
FIWdata <-read_csv("data/FIWdatacsv.csv")
head(FIWdata)
```

## Data cleaning

```{r}
FIWdata <- FIWdata |>
  rename(
    country_territory = `Freedom in the World 2013-2025 Raw Data`,
    region = `...2`,
    ct = `...3`,
    edition = `...4`,
    status = `...5`,
    pr_rating = `...6`,
    cl_rating = `...7`,
    a1 = `...8`, a2 = `...9`, a3 = `...10`, a = `...11`,
    b1 = `...12`, b2 = `...13`, b3 = `...14`, b4 = `...15`, b = `...16`,
    c1 = `...17`, c2 = `...18`, c3 = `...19`, c = `...20`,
    add_q = `...21`, add_a = `...22`, pr = `...23`,
    d1 = `...24`, d2 = `...25`, d3 = `...26`, d4 = `...27`, d = `...28`,
    e1 = `...29`, e2 = `...30`, e3 = `...31`, e = `...32`,
    f1 = `...33`, f2 = `...34`, f3 = `...35`, f4 = `...36`, f = `...37`,
    g1 = `...38`, g2 = `...39`, g3 = `...40`, g4 = `...41`, g = `...42`,
    cl = `...43`, total = `...44`
    )
```

```{r}
FIWdata <- FIWdata[-1, ]
```

## Data visualization

```{r}
FIWdata |>
  mutate(edition = as.numeric(edition), 
         total = as.numeric(total)) |>
  group_by(edition, region) |>
  summarize(mean_total = mean(total, na.rm = TRUE)) |>
  ggplot(aes(x = edition, y = mean_total, color = region)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(2013,2025, by = 1)) +
  labs(
    title = "Freedom in world regions over time",
    x = "Year",
    y = "Total score on freedom-indicating variables",
    caption = "Data from Freedom in the World reports"
  )
```

This scatterplot shows the relationship between region of the world and freedom score over time. Each line represents the mean freedom score of one world region at different points in the last 12 years. 




